Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-12-07 prop_letter_ajhs.tex
prop_letter_ajhs.tex Moved Andy's comments from comment to "Andy says:"

\newcommand{\says}[3]{\begin{framed}\begin{minipage}{0.9\linewidth}\color{#1}{#2 says: #3}\end{minipage}\end{framed}}
\newcommand{\amr}[1]{\says{green}{Amr}{#1}}
\newcommand{\yutsung}[1]{\says{purple}{Yu-Tsung}{#1}}
\newcommand{\gerardo}[1]{\says{OliveGreen}{Gerardo}{#1}}
\newcommand{\andy}[1]{\says{blue}{Andy}{#1}}
% \excludecomment{suggest}
\NewDocumentEnvironment{suggest}{m m}{#1{TEXT \# #2 would go here.}}{#1{TEXT \# #2 end here.}}

@@

We also investigate the second key aspect of imprecise quantum
measurement, which is its statistical nature as reflected in the
application of the Born rule determining the probability of a given
measurement outcome~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}.
Here, Gleason's theorem establishes that there is no alternative to
the Born rule by demonstrating, using reasonable continuity arguments
in Hilbert spaces of dimension $d\ge3$, the existence of a unique
state $\rho$ consistent with the statistical predictions computed from
the Born rule~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.
Our question then concerns what happens when infinite precision cannot
be achieved.  We are able to show that, while a QIVPM incorporating
the effects of finite precision might not be consistent with Gleason's
unique state~$\rho$ on all projectors defined on a Hilbert
space~$\Hilb$ of dimension $d\ge3$, there is a mathematically precise
sense in which one recovers the original Gleason theorem
asymptotically.  Specifically, it is possible to construct a class of
QIVPMs representing bounded resources that is parameterized by the
size of the intervals. We then demonstrate that all QIVPMs in this
class are consistent with a non-empty ``ball'' of quantum states whose
radius is defined by the maximal length of the
intervals characterizing the uncertainties. As the size of the intervals  
goes to zero, this ball of quantum states converges to a
point representing the unique state consistent with the Born rule and 
Gleason's theorem. 
\begin{suggest}{\andy}{2}
We then demonstrate that all QIVPMs in this class are
consistent with non-empty ``balls'' of quantum states whose radii
correspond to the lengths of the intervals characterizing the
uncertainties.  As the sizes of the intervals go to zero, these balls
of quantum states converge to a dense, continuous, set of points
representing the unique states consistent with the Born rule and
Gleason's theorem.
\end{suggest}
%% ajh: above starting at "We then demonstrate" LONG SUGGESTED TEXT #2
%% would go here: use PLURAL, ball -> balls, and then, to me, it would
%% be better as "..converges to a dense set of points."

@@

%% ajh: Inverted logic: LONG SUGGESTED 
\begin{suggest}{\yutsung}{3}
A conventional quantum probability measure can easily be constructed
using the Born rule if one knows the current state of the quantum
system.  Specifically, for each mixed states $\rho =
\sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where $\ket{\phi_j} \in {\cal H}$, $q_j >
0$, and $\sum_{j=1}^{N}q_{j}=1$, the Born rule
induces a probability measure~$\muB_{\rho}$ defined as
$\muB_{\rho}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
\melem{\phi_{j}}{P}{\phi_{j}}$~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}.
\end{suggest}
\begin{suggest}{\andy}{3}
If all the conditions are satisfied, and one knows all the
  $\ket{\phi_j} = \sum_{k=1}^{d}\alpha_{k}\ket{k}$ representing
the current state $\rho$ of the quantum system, then a conventional
probability measure can easily be
constructed~\citep{Born1983bibTeX,Mermin2007,Jaeger2007},  taking the 
form
\[ \muB_{\rho}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}  q_{j}\muB_{\phi_{j}}\left(P\right) \ . \]
\end{suggest}
A conventional quantum probability measure can easily be constructed
using the Born rule if one knows the current state of the quantum
system.  Specifically, for each pure normalized quantum
state~$\ket{\phi} \in {\cal H}$, the Born rule induces a probability measure
$\muB_{\phi}$ defined as
$\muB_{\phi}(P)=\melem{\phi}{P}{\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where $\ket{\phi_j} \in {\cal H}$, $q_j > 0$, and
$\sum_{j=1}^{N}q_{j}=1$, in which case the generalized Born rule
induces a probability measure $\muB_{\rho}$ defined as
$\muB_{\rho}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\muB_{\phi_{j}}\left(P\right)$~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}.

@@

%% ajh: big problems here: LONG SUGGESTED 
\begin{suggest}{\andy}{5}
The above construction assumes a mathematical idealization
having known quantum state and infinitely precise measuring
instruments.  In an actual experimental setup with an ensemble of
quantum states that would ideally be identical, but are not actually
identically prepared, with imperfections and inaccuracies in the
measuring devices,  an experimenter might only be able to determine
that the probability of an event $P$ is concentrated in the range
$[0.49,0.51]$ instead of being precisely $0.5$.   The spread in this
range depends on the amount of resources (time, energy, money, etc.)
that are devoted to the experiment.  In the classical setting, this
``fuzziness'' can be formalized by moving to \emph{interval-valued
  probability measures} (IVPMs), which we explore in the next section,
along with our proposed extension to the quantum domain.
\end{suggest}
%% To
%% begin with, "axioms" is COMPLETELY undefined, appears NOWHERE in
%% the text except here.  So either NAME and LIST (instead of inline
%% text)  the several statements around Eq. (6) as axioms, or use
%% another term.  In addition the english "state of the quantum" is
%% wrong here.  Including fixing that, I would replace the first two
%% sentences by TEXT # 5.  It's really strangely written.
The above axioms describe a mathematical idealization in which the
%%state of the quantum is known and involving infinitely precise
quantum state is known and involving infinitely precise
measurements. In an actual experimental setup with an ensemble of
quantum states whose states are unknown, that are not exactly
identically prepared, and with small imperfections and inaccuracies in
measuring devices, an experimenter might only be able to determine
that the probability of an event $P$ is concentrated in the range
$[0.49,0.51]$ instead of being precisely $0.5$. The spread in this
range depends on the amount of resources (time, energy, money, etc.)
that are devoted to the experiment. In the classical setting, this
``fuzziness'' can be formalized by moving to \emph{interval-valued
  probability measures} (IVPMs), which we explore in the next section,
along with our proposed extension to the quantum domain.

@@

%% ajh: inverted logic and phrasing.  I would start at "To formalize
%% this ..." and replace all the way to Eq (9) by LONG SUGGESTED 
\begin{suggest}{\andy}{6}
We now show that our definition of QIVPMs is consistent
with classical IVPMs in the sense that a restriction of QIVPMs to
mutually commuting events recovers the definition of classical
IVPMs.  However, we must refine this context for the quantum case,
defining a QIVPM  $\bar{\mu}$ to be \emph{consistent} with a state $\rho$ and
projector~$P$ if the interval $\bar{\mu}(P)$ contains the exact
probability calculated by the Born
rule~\citep{Born1983bibTeX,Mermin2007,Jaeger2007} , i.e, if:

@@

%%ajh: This story order is incomprehensible.  In my suggested
%% replacement: LONG SUGGESTED 
\begin{suggest}{\andy}{7}
   The general idea of the proof that QIVPMs become classical
IVPMs when the space of events includes only mutually commuting
events depends on showing the existence of a non-empty core for such
commuting events.  The first step in establishing this assertion is to
to take a subspace $\eventsC$ of mutually commuting events, construct
a partial orthonormal basis by diagonalization, and complete this to a
full orthonormal basis  $\overline{\eventsC}$.  We then
build a bijection between the QIVPM on the set of projectors
associated with this basis and the set of classical events
corresponding to this basis.  Using this correspondence together with a
classical result by
Shapley~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016},
we can establish that for the special case of commuting events, a
QIVPM will always have a non-empty core~\footnote{Shapley's result
  proves that a classical IVPM always contains at least one state that
  is consistent with every event.}.  This leads to
\begin{thm}[Non-empty Core for Compatible Measurements] \label{thm:Shapley}
  For every QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, if a subspace
  of events~$\eventsC\subseteq\events$ commutes, then
  $\coreBorn\left(\bar{\mu},\eventsC\right)\ne\emptyset$,
\end{thm}
whose full proof will appear in ~\cite{TaiThesis2018}.
\end{suggest}
%% I would state FIRST that 'here
%% is the outline of the proof', and LAST 'Thm 1..., whose full proof
%% will appear in [44].
The proof that QIVPMs become classical IVPMs when the space of events
includes only mutually commuting events is presented in detail in a
forthcoming thesis~\cite{TaiThesis2018}. The idea of the proof is as
follows. Given a subspace of mutually commuting events~$\eventsC$, we
can construct a partial orthonormal basis by diagonalization and
complete it to a full orthonormal basis $\overline{\eventsC}$. We then
build a bijection between the QIVPM on the set of projectors
associated with this basis and the set of classical events
corresponding to this basis. Using this correspondence together with a
classical result by
Shapley~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016},
we can establish that for the special case of commuting events, a
QIVPM will always have a non-empty core~\footnote{Shapley's result
  proves that a classical IVPM always contains at least one state that
  is consistent with every event.}.

@@

%% ajh - inverted narrative order, suggested fix in 
%%   LONG SUGGESTED 
\begin{suggest}{\andy}{8}
The importance of the existence of empty cores combined with
  the conditions guaranteeing non-empty cores is that one cannot
  guarantee non-empty cores for finite-precision attempts at proving
  Gleason's theorem by extending the Born measure
  $\muB_{\rho}\left(P\right)$ to QIVPMs $\bar{\mu}\left(P\right)$.  We
  will show in  Sec.~\ref{sec:Gleason} that a core defined by the set
  of states $\coreBorn\left(\bar{\mu},\events'\right)$ that is
  consistent with some arbitrary (not necessarily commuting)
  events~$\events'$  and a particular QIVPM  $\bar{\mu}$
   \emph{may be empty}.
\end{suggest}
The importance of this property is that it is violated in
finite-precision versions of Gleason's theorem
(cf. Thm.~\ref{thm:Non-extensible-of-Gleason's} in Sec.~\ref{sec:Gleason}). As we will show, the set of states
$\coreBorn\left(\bar{\mu},\events'\right)$ that is consistent with
some arbitrary (not necessarily commuting) events~$\events'$ and a
particular QIVPM $\bar{\mu}$ may be \emph{empty}.

>>>>

\newcommand{\says}[3]{\begin{framed}\begin{minipage}{0.9\linewidth}\color{#1}{#2 says: #3}\end{minipage}\end{framed}}
\newcommand{\amr}[1]{\says{green}{Amr}{#1}}
\newcommand{\yutsung}[1]{\says{purple}{Yu-Tsung}{#1}}
\newcommand{\gerardo}[1]{\says{OliveGreen}{Gerardo}{#1}}
\newcommand{\andy}[1]{\says{blue}{Andy}{#1}}
% \excludecomment{suggest}
\NewDocumentEnvironment{suggest}{m m O{}}{#1{TEXT \##2 would go here. #3}}{#1{TEXT \##2 end here.}}

@@

We also investigate the second key aspect of imprecise quantum
measurement, which is its statistical nature as reflected in the
application of the Born rule determining the probability of a given
measurement outcome~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}.
Here, Gleason's theorem establishes that there is no alternative to
the Born rule by demonstrating, using reasonable continuity arguments
in Hilbert spaces of dimension $d\ge3$, the existence of a unique
state $\rho$ consistent with the statistical predictions computed from
the Born rule~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.
Our question then concerns what happens when infinite precision cannot
be achieved.  We are able to show that, while a QIVPM incorporating
the effects of finite precision might not be consistent with Gleason's
unique state~$\rho$ on all projectors defined on a Hilbert
space~$\Hilb$ of dimension $d\ge3$, there is a mathematically precise
sense in which one recovers the original Gleason theorem
asymptotically.  Specifically, it is possible to construct a class of
QIVPMs representing bounded resources that is parameterized by the
size of the intervals. We then demonstrate that all QIVPMs in this
class are consistent with a non-empty ``ball'' of quantum states whose
radius is defined by the maximal length of the
intervals characterizing the uncertainties. As the size of the intervals  
goes to zero, this ball of quantum states converges to a
point representing the unique state consistent with the Born rule and 
Gleason's theorem. 
\begin{suggest}{\andy}{2}[use PLURAL, ball -> balls, and then, to me, it would
be better as "..converges to a dense set of points."]
We then demonstrate that all QIVPMs in this class are
consistent with non-empty ``balls'' of quantum states whose radii
correspond to the lengths of the intervals characterizing the
uncertainties.  As the sizes of the intervals go to zero, these balls
of quantum states converge to a dense, continuous, set of points
representing the unique states consistent with the Born rule and
Gleason's theorem.
\end{suggest}

@@

\begin{suggest}{\yutsung}{3}
A conventional quantum probability measure can easily be constructed
using the Born rule if one knows the current state of the quantum
system.  Specifically, for each mixed states $\rho =
\sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where $\ket{\phi_j} \in {\cal H}$, $q_j >
0$, and $\sum_{j=1}^{N}q_{j}=1$, the Born rule
induces a probability measure~$\muB_{\rho}$ defined as
$\muB_{\rho}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
\melem{\phi_{j}}{P}{\phi_{j}}$~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}.
\end{suggest}
\begin{suggest}{\andy}{3}
If all the conditions are satisfied, and one knows all the
  $\ket{\phi_j} = \sum_{k=1}^{d}\alpha_{k}\ket{k}$ representing
the current state $\rho$ of the quantum system, then a conventional
probability measure can easily be
constructed~\citep{Born1983bibTeX,Mermin2007,Jaeger2007},  taking the 
form
\[ \muB_{\rho}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}  q_{j}\muB_{\phi_{j}}\left(P\right) \ . \]
\end{suggest}
A conventional quantum probability measure can easily be constructed
using the Born rule if one knows the current state of the quantum
system.  Specifically, for each pure normalized quantum
state~$\ket{\phi} \in {\cal H}$, the Born rule induces a probability measure
$\muB_{\phi}$ defined as
$\muB_{\phi}(P)=\melem{\phi}{P}{\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where $\ket{\phi_j} \in {\cal H}$, $q_j > 0$, and
$\sum_{j=1}^{N}q_{j}=1$, in which case the generalized Born rule
induces a probability measure $\muB_{\rho}$ defined as
$\muB_{\rho}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\muB_{\phi_{j}}\left(P\right)$~\citep{Born1983bibTeX,Mermin2007,Jaeger2007}.

@@

\begin{suggest}{\andy}{5}[To
begin with, "axioms" is COMPLETELY undefined, appears NOWHERE in
the text except here.  So either NAME and LIST (instead of inline
text)  the several statements around Eqs.~(\ref{eq:QIVPM-constraints}) as axioms, or use
another term.  In addition the english "state of the quantum" is
wrong here.  Including fixing that, I would replace the first two
sentences by TEXT \# 5.  It's really strangely written.]
The above construction assumes a mathematical idealization
having known quantum state and infinitely precise measuring
instruments.  In an actual experimental setup with an ensemble of
quantum states that would ideally be identical, but are not actually
identically prepared, with imperfections and inaccuracies in the
measuring devices,  an experimenter might only be able to determine
that the probability of an event $P$ is concentrated in the range
$[0.49,0.51]$ instead of being precisely $0.5$.   The spread in this
range depends on the amount of resources (time, energy, money, etc.)
that are devoted to the experiment.  In the classical setting, this
``fuzziness'' can be formalized by moving to \emph{interval-valued
  probability measures} (IVPMs), which we explore in the next section,
along with our proposed extension to the quantum domain.
\end{suggest}

The above axioms describe a mathematical idealization in which the
%%state of the quantum is known and involving infinitely precise
quantum state is known and involving infinitely precise
measurements. In an actual experimental setup with an ensemble of
quantum states whose states are unknown, that are not exactly
identically prepared, and with small imperfections and inaccuracies in
measuring devices, an experimenter might only be able to determine
that the probability of an event $P$ is concentrated in the range
$[0.49,0.51]$ instead of being precisely $0.5$. The spread in this
range depends on the amount of resources (time, energy, money, etc.)
that are devoted to the experiment. In the classical setting, this
``fuzziness'' can be formalized by moving to \emph{interval-valued
  probability measures} (IVPMs), which we explore in the next section,
along with our proposed extension to the quantum domain.

@@

\begin{suggest}{\andy}{6}[inverted logic and phrasing.  I would start at "To formalize
this ..." and replace all the way to Eq.~(\ref{eq:hbar})]
We now show that our definition of QIVPMs is consistent
with classical IVPMs in the sense that a restriction of QIVPMs to
mutually commuting events recovers the definition of classical
IVPMs.  However, we must refine this context for the quantum case,
defining a QIVPM  $\bar{\mu}$ to be \emph{consistent} with a state $\rho$ and
projector~$P$ if the interval $\bar{\mu}(P)$ contains the exact
probability calculated by the Born
rule~\citep{Born1983bibTeX,Mermin2007,Jaeger2007} , i.e, if:

@@

\begin{suggest}{\andy}{7}[This story order is incomprehensible. I would state FIRST that 'here
is the outline of the proof', and LAST 'Thm 1..., whose full proof
will appear in~\cite{TaiThesis2018}.]
   The general idea of the proof that QIVPMs become classical
IVPMs when the space of events includes only mutually commuting
events depends on showing the existence of a non-empty core for such
commuting events.  The first step in establishing this assertion is to
to take a subspace $\eventsC$ of mutually commuting events, construct
a partial orthonormal basis by diagonalization, and complete this to a
full orthonormal basis  $\overline{\eventsC}$.  We then
build a bijection between the QIVPM on the set of projectors
associated with this basis and the set of classical events
corresponding to this basis.  Using this correspondence together with a
classical result by
Shapley~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016},
we can establish that for the special case of commuting events, a
QIVPM will always have a non-empty core~\footnote{Shapley's result
  proves that a classical IVPM always contains at least one state that
  is consistent with every event.}.  This leads to
\begin{thm}[Non-empty Core for Compatible Measurements] \label{thm:Shapley}
  For every QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, if a subspace
  of events~$\eventsC\subseteq\events$ commutes, then
  $\coreBorn\left(\bar{\mu},\eventsC\right)\ne\emptyset$,
\end{thm}
whose full proof will appear in ~\cite{TaiThesis2018}.
\end{suggest}
The proof that QIVPMs become classical IVPMs when the space of events
includes only mutually commuting events is presented in detail in a
forthcoming thesis~\cite{TaiThesis2018}. The idea of the proof is as
follows. Given a subspace of mutually commuting events~$\eventsC$, we
can construct a partial orthonormal basis by diagonalization and
complete it to a full orthonormal basis $\overline{\eventsC}$. We then
build a bijection between the QIVPM on the set of projectors
associated with this basis and the set of classical events
corresponding to this basis. Using this correspondence together with a
classical result by
Shapley~\cite{Shapley1971,GilboaSchmeidler1994,NgMoYeh1997,Grabisch2016},
we can establish that for the special case of commuting events, a
QIVPM will always have a non-empty core~\footnote{Shapley's result
  proves that a classical IVPM always contains at least one state that
  is consistent with every event.}.

@@

\begin{suggest}{\andy}{8}[inverted narrative order]
The importance of the existence of empty cores combined with
  the conditions guaranteeing non-empty cores is that one cannot
  guarantee non-empty cores for finite-precision attempts at proving
  Gleason's theorem by extending the Born measure
  $\muB_{\rho}\left(P\right)$ to QIVPMs $\bar{\mu}\left(P\right)$.  We
  will show in  Sec.~\ref{sec:Gleason} that a core defined by the set
  of states $\coreBorn\left(\bar{\mu},\events'\right)$ that is
  consistent with some arbitrary (not necessarily commuting)
  events~$\events'$  and a particular QIVPM  $\bar{\mu}$
   \emph{may be empty}.
\end{suggest}
The importance of this property is that it is violated in
finite-precision versions of Gleason's theorem
(cf. Thm.~\ref{thm:Non-extensible-of-Gleason's} in Sec.~\ref{sec:Gleason}). As we will show, the set of states
$\coreBorn\left(\bar{\mu},\events'\right)$ that is consistent with
some arbitrary (not necessarily commuting) events~$\events'$ and a
particular QIVPM $\bar{\mu}$ may be \emph{empty}.
